Evolution, functional diversity, regime shifts, hysteresis, and nonlinearity; ecosystem responses to environmental change.

### 2021-11-25 experiments/find discont/find discont.Rmd
added temporal SS finding, and reorganise

---
title: "3 strains"
author: "Owen Petchey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
  html_document:
    keep_md: yes
---

@@

# Temporal switching

```{r}
## Here we get some rows of var_expt that correspond with particular diversity levels

var_expt_levels <- var_expt[, 1:6]

no_diversity <- which(rowSums(abs(var_expt_levels)) == 0)
max_diversty_all <- which(
  max(rowSums(abs(var_expt_levels))) == rowSums(abs(var_expt_levels))
)
max_only_CB_diversity <- which(
  max(rowSums(abs(var_expt_levels[, 1:2]))) == rowSums(abs(var_expt_levels[, 1:2])) &
  rowSums(abs(var_expt_levels[,3:6])) == 0
)
# var_expt_levels[381,]

max_only_SBPB_diversity <- which(
  max(rowSums(abs(var_expt_levels[, 3:6]))) == rowSums(abs(var_expt_levels[, 3:6])) &
  rowSums(abs(var_expt_levels[, 1:2])) == 0
)
#var_expt_levels[20,]

medium_diverity_varexp_row <- 311

@@

## Oxic to anoxic

### No diversity

```{r eval = eval_dynamics_flag}
parameter$sim_duration <- 80000
```

```{r eval = eval_dynamics_flag}
parameter$log10a_series <- c(-1, -7, -7)
parameter$strain_parameter <- var_expt$pars[[no_diversity]]
parameter$strain_parameter$initial_state <- new_initial_state(
  num_CB_strains,
  num_PB_strains,
  num_SB_strains,
  values = "bush_ssfig3"
)
parameter$strain_parameter$initial_state[grep("CB_", names(parameter$strain_parameter$initial_state))] <- 10^10/num_CB_strains
sim_res_novar1 <- run_simulation(parameter)
saveRDS(sim_res_novar1, here("experiments/3_strains/data/sim_res_novar1.RDS"))
```

```{r}
sim_res_novar1 <- readRDS(here("experiments/3_strains/data/sim_res_novar1.RDS"))
plot_dynamics(sim_res_novar1)
#ggsave(here("simulations/expt2/figures/switching_novar.pdf"), width = 10)
```

### With diversity

```{r eval = eval_dynamics_flag}
parameter$strain_parameter <- var_expt$pars[[medium_diverity_varexp_row]]
parameter$strain_parameter$initial_state <- sim_res_novar1$strain_parameter$initial_state
sim_res_highvar1 <- run_simulation(parameter)
saveRDS(sim_res_highvar1, here("experiments/3_strains/data/sim_res_highvar1.RDS"))
```

```{r}
sim_res_highvar1 <- readRDS(here("experiments/3_strains/data/sim_res_highvar1.RDS"))
plot_dynamics(sim_res_highvar1)
#ggsave(here("simulationsexpt2/figures/switching_highvar.pdf"), width = 10)

## Anoxic to oxic

### No diversity

```{r eval = eval_dynamics_flag}
parameter$sim_duration <- 60000
```

```{r eval = eval_dynamics_flag}
parameter$log10a_series <- c(-5, -3, -1, -1)
parameter$strain_parameter <- var_expt$pars[[no_diversity]]
parameter$strain_parameter$initial_state <- new_initial_state(
  num_CB_strains,
  num_PB_strains,
  num_SB_strains,
  values = "bush_ssfig3"
)
parameter$strain_parameter$initial_state[grep("CB_", names(parameter$strain_parameter$initial_state))] <- 10/num_CB_strains
sim_res_novar2 <- run_simulation(parameter)
saveRDS(sim_res_novar2, here("experiments/3_strains/data/sim_res_novar2.RDS"))
```

```{r}
sim_res_novar2 <- readRDS(here("experiments/3_strains/data/sim_res_novar2.RDS"))
plot_dynamics(sim_res_novar2)
#ggsave(here("simulations/expt2/figures/switching_novar.pdf"), width = 10)
```

### With diversity

```{r eval = eval_dynamics_flag}
parameter$strain_parameter <- var_expt$pars[[medium_diverity_varexp_row]]
parameter$strain_parameter$initial_state <- sim_res_novar2$strain_parameter$initial_state
sim_res_highvar2 <- run_simulation(parameter)
saveRDS(sim_res_highvar2, here("experiments/3_strains/data/sim_res_highvar2.RDS"))
```

```{r}
sim_res_highvar2 <- readRDS(here("experiments/3_strains/data/sim_res_highvar2.RDS"))
plot_dynamics(sim_res_highvar2)
#ggsave(here("simulationsexpt2/figures/switching_highvar.pdf"), width = 10)

## Anoxic to oxic to anoxic

### No diversity

```{r eval = eval_dynamics_flag}
parameter$sim_duration <- 1000000
```

```{r eval = eval_dynamics_flag}
parameter$minimum_abundances <- rep(100, 3)
names(parameter$minimum_abundances) <- c("CB", "PB", "SB")
```

```{r eval = eval_dynamics_flag}
parameter$log10a_series <- c(-1, -8, -1)
parameter$strain_parameter <- var_expt$pars[[no_diversity]]
parameter$strain_parameter$initial_state <- new_initial_state(
  num_CB_strains,
  num_PB_strains,
  num_SB_strains,
  values = "bush_ssfig3"
)
parameter$strain_parameter$initial_state[grep("CB_", names(parameter$strain_parameter$initial_state))] <- 10/num_CB_strains
sim_res_novar3 <- run_simulation(parameter)
saveRDS(sim_res_novar3, here("experiments/3_strains/data/sim_res_novar3.RDS"))
```

```{r}
sim_res_novar3 <- readRDS(here("experiments/3_strains/data/sim_res_novar3.RDS"))
plot_dynamics(sim_res_novar3)
#ggsave(here("simulations/expt2/figures/switching_novar.pdf"), width = 10)
```

### With diversity

```{r eval = eval_dynamics_flag}
parameter$strain_parameter <- var_expt$pars[[medium_diverity_varexp_row]]
parameter$strain_parameter$initial_state <- sim_res_novar3$strain_parameter$initial_state
sim_res_highvar3 <- run_simulation(parameter)
saveRDS(sim_res_highvar3, here("experiments/3_strains/data/sim_res_highvar3.RDS"))
```

```{r}
sim_res_highvar3 <- readRDS(here("experiments/3_strains/data/sim_res_highvar3.RDS"))
plot_dynamics(sim_res_highvar3)
#ggsave(here("simulationsexpt2/figures/switching_highvar.pdf"), width = 10)

### Ecosystem-environment relationship

```{r eval = FALSE, echo = FALSE}
visualise_temporal_env_eco(sim_res_novar3, sim_res_highvar3)
```

# Stable state finding

## Finding

### Setup parameter

```{r}
options(mc.cores = 7)
```

```{r}
minimum_abundances <- rep(0, 3)
names(minimum_abundances) <- c("CB", "PB", "SB")

grid_num_a <- 1000 #usually 1000 ## number of a_0 values
#grid_num_a <- 10 ## FOR TEST
a_Os <- 10^seq(-7, -0.5, length=grid_num_a) ## sequence of a_0 values
grid_num_N <- 2 ## number of N values
initial_CBs <- 10^seq(0, 10, length=grid_num_N) ## sequence of N values
initial_PBs <- 1e8 ## not varied
initial_SBs <- 1e8 ## not varied
# next line creates all possible combinations
ss_expt <- expand.grid(N_CB = initial_CBs,
                      N_PB = initial_PBs,
                      N_SB = initial_SBs,
                      a_O = a_Os)

parameter <- new_ss_by_a_N_parameter(
  dynamic_model = parameter$dynamic_model,
  event_definition = parameter$event_definition,
  event_interval = 1000000,
  noise_sigma = parameter$noise_sigma,
  minimum_abundances = minimum_abundances,
  sim_duration = 1000000,
  sim_sample_interval = 1000000,
  log10a_series = parameter$log10a_series,
  solver_method = parameter$solver_method,
  ss_expt = ss_expt
)
rm(minimum_abundances, grid_num_a, a_Os, grid_num_N, initial_CBs, initial_PBs, initial_SBs, ss_expt)

saveRDS(parameter, here("experiments/3_strains/data/parameter_1e6_x2x6_factorial.RDS"))
saveRDS(var_expt, here("experiments/3_strains/data/var_expt_1e6_x2x6_factorial.RDS"))
```

### Run stable state finding

*Careful, this simulation takes about 600 hours on a single core

```{r run_experiment, eval = eval_dynamics_flag}
run_ss_var_experiment(
  parameter = readRDS(here("experiments/3_strains/data/parameter_1e6_x2x6_factorial.RDS")), 
  var_expt = readRDS(here("experiments/3_strains/data/var_expt_1e6_x2x6_factorial.RDS"))) %>%
saveRDS(here("experiments/3_strains/data/ss_data_1e6_x2x6_factorial.RDS"))
```

### Process the stable state data

Bring in various stable state datasets
```{r eval = eval_dynamics_flag}
cluster <- multidplyr::new_cluster(7)
multidplyr::cluster_library(cluster, c("microxanox", "dplyr"))

## sim length 1'000'000, 20 SBPBgrad, 2xCB variation, 6xSBPB variation
readRDS(here("experiments/3_strains/data/ss_data_1e6_x2x6_factorial_small.RDS")) %>%
  mutate(sim_length = 1000000) %>% 
  multidplyr::partition(cluster) %>%
  mutate(stability_measures = list(get_stability_measures(ss_res))) %>%
  collect() %>%
  unnest(cols = c(stability_measures)) %>%
  saveRDS(here("experiments/3_strains/data/stab_data_1e6_x2x6_factorial_small.RDS"))
```

```{r}
## find various combinations of diversity
var_expt <- readRDS(here("experiments/3_strains/data/ss_data_1e6_x2x6_factorial.RDS"))
stab_data <- readRDS(here("experiments/3_strains/data/stab_data_1e6_x2x6_factorial.RDS"))

var_expt_levels <- var_expt[,1:6]

no_diversity <- which(rowSums(abs(var_expt_levels))==0)
max_diversty_all <- which(max(rowSums(abs(var_expt_levels))) ==
                            rowSums(abs(var_expt_levels)))
max_only_CB_diversity <- which(max(rowSums(abs(var_expt_levels[,1:2]))) ==
                            rowSums(abs(var_expt_levels[,1:2])) &
                              rowSums(abs(var_expt_levels[,3:6]))==0)
#var_expt_levels[381,]

max_only_SBPB_diversity <- which(max(rowSums(abs(var_expt_levels[,3:6]))) ==
                            rowSums(abs(var_expt_levels[,3:6])) &
                              rowSums(abs(var_expt_levels[,1:2]))==0)
#var_expt_levels[20,]

medium_diverity_varexp_row <- 311

### Display results

```{r eval = plot_ss_results}

p1  <- plot_ss_result1(var_expt,
                result_index = no_diversity,
                filename_prefix = NULL,
                save_image_file = FALSE)
p1

p2  <- plot_ss_result1(var_expt,
                result_index = max_diversty_all,
                filename_prefix = NULL,
                save_image_file = FALSE)
p2

p3  <- plot_ss_result1(var_expt,
                result_index = max_only_CB_diversity,
                filename_prefix = NULL,
                save_image_file = FALSE)
p3

p4  <- plot_ss_result1(var_expt,
                result_index = max_only_SBPB_diversity,
                filename_prefix = NULL,
                save_image_file = FALSE)
p4

p_overlay1 <- plot_ss_result2(var_expt[no_diversity,]$ss_res[[1]],
                             var_expt[max_diversty_all,]$ss_res[[1]],
                             xlims = c(-7, -1))
p_overlay1

p_overlay2 <- plot_ss_result2(var_expt[no_diversity,]$ss_res[[1]],
                             var_expt[max_only_CB_diversity,]$ss_res[[1]],
                             xlims = c(-7, -1))
p_overlay2

p_overlay3 <- plot_ss_result2(var_expt[no_diversity,]$ss_res[[1]],
                             var_expt[max_only_SBPB_diversity,]$ss_res[[1]],
                             xlims = c(-7, -1))
p_overlay3

```{r eval = plot_ss_results}
tempp1 <- var_expt[no_diversity,]$ss_res[[1]] %>%
  filter(initial_N_CB == 1e10,
         log10(a_O) > -4.5, log10(a_O) < -4)
tempp2 <- var_expt[max_only_SBPB_diversity,]$ss_res[[1]] %>%
  filter(initial_N_CB == 1e10,
         log10(a_O) > -4.5, log10(a_O) < -4)

ggplot() +
  geom_line(mapping = aes(x = log10(tempp1$a_O),
                          y = log10(tempp1$O))) +
  geom_line(mapping = aes(x = log10(tempp2$a_O),
                          y = log10(tempp2$O)),
            col = "blue")

ggplot() +
  geom_histogram(mapping = aes(x = log10(tempp1$O) - log10(tempp2$O)),
            col = "blue")

ggplot() +
  geom_line(mapping = aes(x = log10(tempp1$a_O),
    y = log10(tempp1$P) - log10(tempp2$P)),
            col = "blue")

```{r eval = plot_ss_results}
CB_vars <- unique(stab_data$CB_var_gmax_s)
SB_vars <- unique(stab_data$SB_var_gmax_s)

CB_stab_data <- stab_data %>%
  filter(SB_var_gmax_s == 0) %>%
  mutate(var_treat = "CB",
         var_gmax = CB_var_gmax_s)

SBPB_stab_data <- stab_data %>%
  filter(CB_var_gmax_s == 0) %>%
  mutate(var_treat = "SB-PB",
         var_gmax = SB_var_gmax_s)

for_join <- tibble(CB_var_gmax_s = CB_vars,
                     SB_var_gmax_s = SB_vars)
CBSBPB_stab_data <- stab_data %>%
  right_join(for_join) %>%
  mutate(var_treat = "CB-SB-PB",
         var_gmax = CB_var_gmax_s)

all_stab_results <- CB_stab_data %>%
  bind_rows(SBPB_stab_data) %>%
#  bind_rows(results3) %>%
 # bind_rows(results4) %>%
  bind_rows(CBSBPB_stab_data)

all_stab_results <- all_stab_results %>%
    mutate(var_treat = forcats::fct_relevel(var_treat, levels = c("CB", "SB-PB", "CB-SB-PB")))

#saveRDS(all_stab_results, here("experiments/experiment summary/all_stab.RDS"))

#all_stab_results <- readRDS(here("experiments/experiment summary/all_stab.RDS"))
```

### Raw
```{r eval = plot_ss_results}
all_stab_results %>%
  filter(Species == "O") %>%
  ggplot(aes(x = var_gmax, y = hyst_range_raw, col=var_treat, shape = as_factor(sim_length))) +
  geom_point() +
  xlab("Amount of trait variation\n[see text for units]") +
  ylab("Extent of bistability region\n[log10 oxygen diffusivity]") +
  labs(col = "Variation in\nonly these\nfunctional groups")
##ggsave("manuscript/figures/extent_of_bistab1.pdf", height = 4)

all_stab_results %>%
  #filter(var_treat == "CB") %>%
  filter(Species == "O",
         sim_length == 1e6) %>%
  ggplot(aes(x = var_gmax,
             ymin = hyst_min_raw,
             ymax = hyst_max_raw,
             fill=var_treat)) +
  geom_ribbon(alpha = 0.5) +
  facet_wrap( ~ var_treat, nrow = 3) +
  xlab("Amount of trait variation\n[see text for units]") +
  ylab("Oxygen diffusivity\n[log10 uM per hour]") +
  labs(fill = "Variation in\nonly these\nfunctional groups") +
  coord_flip() +
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank()
  )

##ggsave("manuscript/figures/extent_of_bistab2.pdf", height = 4)
stab_data %>%
  filter(Species == "O",
         sim_length == 1e6) %>%
  ggplot(aes(x = CB_var_gmax_s, y = SB_var_gmax_s, fill = hyst_range_raw, col = hyst_range_raw)) +
  geom_point(size = 3)

##ggsave("manuscript/figures/extent_of_bistab2.pdf", height = 4)

stab_data %>%
  filter(Species == "O",
         sim_length == 1e6) %>%
  ggplot(aes(x = CB_var_gmax_s, y = hyst_range_raw, col = as.factor(SB_var_gmax_s))) +
  geom_line(size = 2)
```

### Log transformed

```{r eval = plot_ss_results}
all_stab_results %>%
  filter(Species == "O") %>%
  ggplot(aes(x = var_gmax, y = hyst_range_log, col=var_treat, shape = as_factor(sim_length))) +
  geom_point() +
  xlab("Amount of trait variation\n[see text for units]") +
  ylab("Extent of bistability region\n[log10 oxygen diffusivity]") +
  labs(col = "Variation in\nonly these\nfunctional groups")
##ggsave("manuscript/figures/extent_of_bistab1.pdf", height = 4)

all_stab_results %>%
  #filter(var_treat == "CB") %>%
  filter(Species == "O",
         sim_length == 1e6) %>%
  ggplot(aes(x = var_gmax,
             ymin = hyst_min_log,
             ymax = hyst_max_log,
             fill=var_treat)) +
  geom_ribbon(alpha = 0.5) +
  facet_wrap( ~ var_treat, nrow = 3) +
  xlab("Amount of trait variation\n[see text for units]") +
  ylab("Oxygen diffusivity\n[log10 uM per hour]") +
  labs(fill = "Variation in\nonly these\nfunctional groups") +
  coord_flip() +
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank()
  )

##ggsave("manuscript/figures/extent_of_bistab2.pdf", height = 4)
stab_data %>%
  filter(Species == "O",
         sim_length == 1e6) %>%
  ggplot(aes(x = CB_var_gmax_s, y = SB_var_gmax_s, fill = hyst_range_log, col = hyst_range_log)) +
  geom_point(size = 3)

##ggsave("manuscript/figures/extent_of_bistab2.pdf", height = 4)

stab_data %>%
  filter(Species == "O",
         sim_length == 1e6) %>%
  ggplot(aes(x = CB_var_gmax_s, y = hyst_range_log, col = as.factor(SB_var_gmax_s))) +
  geom_line(size = 2)
```

>>>>

---
title: "find discont"
author: "Owen Petchey"
date: "11/15/2021"
output: html_document
---
---
title: "3 strains"
author: "Owen Petchey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
  html_document:
    keep_md: yes
---

@@

```{r eval = eval_dynamics_flag}
parameter$sim_duration <- 8000
parameter$sim_sample_interval <- parameter$sim_duration
parameter$log10a_series <- c(-4, -4)
parameter$strain_parameter <- var_expt$pars[[no_diversity]]
parameter$strain_parameter$initial_state <- new_initial_state(
  num_CB_strains,
  num_PB_strains,
  num_SB_strains,
  values = "bush_ssfig3"
)

get_final_oxygen <- function(x) {
  parameter$strain_parameter$initial_state[grep("CB_", names(parameter$strain_parameter$initial_state))] <- 10 ^ (x) /
    num_CB_strains
  sim_res_novar1 <- run_simulation(parameter)
  temp <- sim_res_novar1$result$O
  temp[length(temp)]
}
get_final_oxygen(1)
get_final_oxygen(2)
get_final_oxygen(3)
get_final_oxygen(4)
get_final_oxygen(5)
get_final_oxygen(6)
get_final_oxygen(7)
get_final_oxygen(8)
get_final_oxygen(9)
get_final_oxygen(10)

parameter$strain_parameter$initial_state[grep("CB_", names(parameter$strain_parameter$initial_state))] <- 10^1/num_CB_strains
sim_res_novar1 <- run_simulation(parameter)
sim_res_novar1$result$O

@@

### 2021-11-25 manuscript/article.Rmd
added temporal SS finding, and reorganise

The model ecosystem contains three types (functional groups) of organism (Figure \ref(fig:ecosys_network)): cyanobacteria ($CB$), sulfate-reducing bacteria ($SB$), and phototrophic sulfur bacteria ($PB$), four chemical substrates (oxygen $O$, phosphorous $P$, reduced sulfur $SR$, and oxidised sulfur $SO$), and four types of flow/interaction (production, consumption, inhibition, and diffusion). Particularly important among these interactions are that the cyanobacteria produce oxygen which inhibits the growth of both types of sulfur bacteria. And the sulfate-reducing bacteria produce reduced sulfur, such as hydrogen sulphide, which inhibit the growth of cyanobacteria. This mutual inhibition (a positive feedback) creates the potential for an oxic cyanobacteria dominated state and an anoxic sulfur bacteria dominated state to be alternate stable states.

>>>>

The model ecosystem contains three types (functional groups) of organism (Figure \@ref(fig:ecosys_network)): cyanobacteria ($CB$), sulfate-reducing bacteria ($SB$), and phototrophic sulfur bacteria ($PB$), four chemical substrates (oxygen $O$, phosphorous $P$, reduced sulfur $SR$, and oxidised sulfur $SO$), and four types of flow/interaction (production, consumption, inhibition, and diffusion). Particularly important among these interactions are that the cyanobacteria produce oxygen which inhibits the growth of both types of sulfur bacteria. And the sulfate-reducing bacteria produce reduced sulfur, such as hydrogen sulphide, which inhibit the growth of cyanobacteria. This mutual inhibition (a positive feedback) creates the potential for an oxic cyanobacteria dominated state and an anoxic sulfur bacteria dominated state to be alternate stable states.
